---
title: "Environmental Niche Modelling in R - Data Cleaning"
author: "Jack Richardson u7291420"
date: "4/10/23"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
editor_options: 
  markdown: 
    wrap: 72
---

# Platypus Data

```{r}
# Read Platypus Data
raw_platypus <- read.csv("raw_data/Platypus18-10-23.csv")
```

# Rivers Data

```{r}
# Find Rivers Layer
ogrListLayers("raw_data/rivers_australia_37252")

# Read Rivers Data
raw_rivers <- sf::st_read(dsn = "raw_data/rivers_australia_37252", layer = "rivers_australia_37252")
```

# 

## Cleaning Platypus Data
Clean platypus Data
```{r}
# Remove duplicates
  # We will assume that if the date, location and recorder is the same, then the result is a duplicate
  # The Location change should prevent filtering out multiple sightings in one day
cleaning_platypus <- raw_platypus %>%
  distinct(eventDate, decimalLatitude, decimalLongitude, recordedBy, .keep_all = TRUE)
  # From 20077
  # To 18662
  # Removed 1415 results

# Remove absence
cleaning_platypus <- filter(cleaning_platypus, occurrenceStatus == "PRESENT")

# Remove Location Uncertianty over 10,000m
cleaning_platypus <- filter(cleaning_platypus, coordinateUncertaintyInMeters <= 10000)

#Finalise cleaning

write.csv(cleaning_platypus, "processed_data/platypus.csv")
platypus <- cleaning_platypus
```

## Cleaning River Data
Clean Rivers Data

The following areas lie significantly outside of platypus habitats, and will thus be filtered from data to lessen computational load:

- Australia, North Coast
- Australia, West Coast
- New Zealand
- South Pacific Islands

Australia, interior slightly overlaps with some recorded sightings of platypus, so should be included in the dataset. Removing rivers will not effect the model

```{r}
ausRegionKeep <- c(
  "Australia, East Coast", 
  "Australia, South Coast",
  "Australia, Interior", 
  "Murray - Darling", 
  "Tasmania")

rivers <- filter(raw_rivers, MAJ_NAME == ausRegionKeep)

# Create blank dataframe to add data to
SortedRivers <- data.frame(matrix(ncol = 5, nrow = 0, dimnames = list(NULL, c("ARCID", "decimalLongitude", "decimalLatitude", "MAJ_NAME", "SUB_NAME"))))


# Isolate the river geometry (list form of each point of river)
geometry_column <- rivers$geometry

# Extracting the point data
# Note. This will take up to 10 minutes to complete.
for (x in 1:nrow(rivers)) {

  # Unnest coordinates from list
  coordinates <- st_coordinates(geometry_column[x])

  for (y in 1:nrow(coordinates)) {
    tempDF <- data.frame(
      "ARCID" = as.integer(rivers[[x, "ARCID"]]),
      "decimalLongitude" = coordinates[y, "Y"],
      "decimalLatitude" = coordinates[y, "X"],
      "MAJ_NAME" = rivers[[x, "MAJ_NAME"]],
      "SUB_NAME" = rivers[[x, "SUB_NAME"]])

    # use rbind() to add row to bottom of dataframe
  SortedRivers <- rbind(SortedRivers, tempDF)

  # Uncomment below code to see progress
  ## cat("Progress:   ", x, " of " nrow(rivers), "      ")
  }
}

write.csv(SortedRivers, "processed_data/SortedRivers.csv")

ggplot(SortedRivers, aes(x = decimalLongitude, y = decimalLatitude, colour = MAJ_NAME)) +
  geom_point()
```

We can clearly see a large area of Australia, especially platypus inhabited East coast, is populated with rivers within our dataset. Whilst it may not contain every stream and waterway, it will do for the project