---
title: "Environmental Niche Modelling in R"
author: "Jack Richardson u7291420"
date: "4/10/23"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
editor_options: 
  markdown: 
    wrap: 72
---

# Word/figure count

Words: [The number of words in your document, calculated using the
word_count() function at the end of this document] Figures: 7

# Location on GitHub

<https://github.com/JackRicho/DS4B-final-project>

# Introduction

## What is ENM? and why BioClim?
ENM (Environmental Niche Modeling), commonly referred to as Species Distribution Modelling (SDM) is the use of computational models to calculate habitat suitability for a particular species, and thus where they may be found. 

There are a variety of approaches a model can take I have selected to use BioClim, which inputs geographic observation data of platypus, then finds key relations of presence and absence with a number of climatic variables [Booth et.al, 2013]. The model then gives each square on a grid map of Australia a score, whereby which higher values indicate more suitable conditions, and lower values represent less suitable conditions.

BioClim is not the only approach to SDM. More complex methods, such as Generalized Linear Modelling, and less complex methods, such as Domain, are available. Additionally, alternative approaches exist, like Random Forests, which is based on decision trees rather than observation data, making it, in theory, possible to model species of which limited observations exist, such as introduced species. 

For this project, BioClim is the most appropriate option, as it is designed not to predict places of new sightings, but to calculate the relative suitability of habitats, assisting in identifying areas of likely observation.

## Why Platypus?
Platypus are nocturnal aquatic mammals (monotremes) that spend large amounts of time within burrows accessible only underground and underwater [Biano et. al, 2019]. As such, the best time to see them is nearly exclusively at night (or early mornings and late evenings in winter). To add to this, they are beautiful creatures and some of the most morphologically unique on the world. It has been my project for the year to spot one in the wild, and ENM can help do so.


# Data Description
This project uses 3 datasets:

Atlas of Living Australia's (ALA) most current (as of 18/10/22) platypus
occurrence records with no exclusions to data. Collected through a
variety of research, conservation work and citizen science

Bioclimatic environment data that represents annual trends in climate,
rainfall and seasonalty. Standard set of variables used in Species
Distribution Modeling

Rivers Dataset containing geospatial point locations of Australia and
New Zealand Rivers derived from World Wildlife Fund's (WWF) HydroSHEDS
drainage direction layer and a stream network layer.

# Questions/Aims

This project has 2 aims:

-   Primary Aim: To create a species distribution model of Platypus using the bioclim
    model.

-   Secondary Aim: To create a tool whereby users can input a location and they are
    told the suitability of that place to platypus habitation, and are
    told a nearby river where they may be able to find platypus
    



```{r}
library(sf)
library(rgdal)
library(dplyr)
library(devtools)
library(ENMTools)
library(tidyverse)
library(leaflet)
library(terra)
library(rnaturalearth)
```

# Raw data

## Platypus Observation Data

Platypus Data: Available through Atlas of Living Australia. Search:
*Ornithorhynchus anatinus*

available here:
<https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/ac61fd14-4950-4566-b384-304bd99ca75f>

acquired 18-10-2023 with no data exclusions. Renamed
(Speciescommonname)DD-MM-YY i.e. Platypus18-10-23

## Rivers of Australia & NZ Data

Rivers Data: Available through Food and Agriculture Organization of the
United Nations: /AQUASTAT/Rivers of Australia and New Zealand In ERSI
data format (geospatial data)

available here:
<https://data.apps.fao.org/catalog/iso/6a53d768-1e20-46ea-92a8-c4040286057d>

## Environmental Variables and BioClim Raster Data

Environment Data: Environment data available through raster package

```{r}
# Get Environment Data
env <- raster::getData(name = "worldclim", res = 10, var = "bio")
```

# Data wrangling
## Rivers Data

Data cleaning can be found on the github linked at the top, or here:
[https://github.com/JackRicho/DS4B-final-project/blob/main/data_cleaning.html]

From its raw form, the rivers dataset was extracted so that each point, originally in list form, is presented in a table with its river name attatched to it
```{r}
SortedRivers <- read.csv("processed_data/SortedRivers.csv")
```

## Platypus Data
Data was cleaned to remove high location uncertianty values, duplicate entries, and absence records.
```{r}
platypus <- read.csv("processed_data/platypus.csv")
```

## Wrangling Environmental Variable Data

Environmental data
```{r}
# Trim down the environmental layer to frame data better, include Australia only
env <- crop(env, extent(110, 155, -45, -8))
```


```{r include=FALSE}
bioclimVariables <-data.frame(
  BIO = c(
    "BIO1", "BIO2", "BIO3", "BIO4", "BIO5", "BIO6", "BIO7", "BIO8", "BIO9", 
    "BIO10", "BIO11", "BIO12", "BIO13", "BIO14", "BIO15", "BIO16", "BIO17", 
    "BIO18", "BIO19"),
  Name = c(
    "Annual Mean Temperature (C*100)", 
    "Mean Diurnal Range (Mean of monthly (max temp - min temp)) (C*100)", 
    "Isothermality (BIO2/BIO7) (×100)", 
    "Temperature Seasonality (standard deviation ×100)", 
    "Max Temperature of Warmest Month (C*100)", 
    "Min Temperature of Coldest Month (C*100)", 
    "Temperature Annual Range (BIO5-BIO6) (C*100)", 
    "Mean Temperature of Wettest Quarter (C*100)", 
    "Mean Temperature of Driest Quarter (C*100)", 
    "Mean Temperature of Warmest Quarter (C*100)", 
    "Mean Temperature of Coldest Quarter (C*100)", 
    "Annual Precipitation (mm)", 
    "Precipitation of Wettest Month (mm)", 
    "Precipitation of Driest Month (mm)", 
    "Precipitation Seasonality (Coefficient of Variation) (mm)", 
    "Precipitation of Wettest Quarter (mm)", 
    "Precipitation of Driest Quarter (mm)", 
    "Precipitation of Warmest Quarter (mm)", 
    "Precipitation of Coldest Quarter (mm)")
)
```


```{r}
print(bioclimVariables)
```

# Sanity checks

## Environment Sanity Checks

```{r}
# Check Environment Data
plot(env[[1]], main = bioclimVariables$Name[1])
  # Framing is appropriate
```

## Platypus Sanity Checks

```{r}
# Use naturalEarth library to give a map outline of Australia
australia <- ne_countries(country = "Australia", returnclass = "sf")

# Create blank map of Australia
map <- ggplot() +
  geom_sf(data = australia, fill = "lightblue") +
  theme_void()

# Plot our platypus observations onto NaturalEarth map
map_with_points <- map +
  geom_point(data = platypus, aes(x = decimalLongitude, y = decimalLatitude), size = 0.1) +
  ggtitle("Platypus Observations from ALA")

print(map_with_points)
```

## River Coordinate Point Extraction & Sanity Checks

```{r}
# Plot points to ensure large amount of rivers in Australia contained within dataset
# And overlay platypus sightings to ensure most areas are located close to river

# Reuse NaturalEarth map
rivers_map_with_points <- map +
  geom_point(data = SortedRivers, aes(x = decimalLongitude, y = decimalLatitude), size = 0.1) +
  geom_point(data = platypus, aes(x = decimalLongitude, y = decimalLatitude), size = 0.1, colour = 'red') 

print(rivers_map_with_points)
```
Red points are observed platypus sightings from ALA. Black points are coordinates along the rivers.

# Addressing the questions/aims

## Creating a bioclim model

```{r}
# Create DF with only required data
platypusPresenceDF <- platypus %>% 
  dplyr::select(decimalLongitude, decimalLatitude)
  # I have attempted to create this in the SpatRaster format, however it encounters issues down the line. Will work fine


# We package each lineage up in an ENMtools species object
# Creates empty object with slots for various values
platypusENM <- enmtools.species()

#Inhabit columns with required data
platypusENM$species.name <- "Ornithorhynchus anatinus"
platypusENM$presence.points <- platypusPresenceDF

# Check format of the enmtools.species
platypusENM <- check.species(platypusENM)
  # Populates object with required missing data
```

## BioClim Model

```{r}
# Use enmTools to create a bioclim model
platypus.bc <- enmtools.bc(platypusENM, env, test.prop = 0.3)
  # Bioclim uses the presence of observed instances and analyses them in relation to our 19 climatic variables to create a suitability plot
```


### Response Plots

Understanding Bioclim Plots:

The x axis represents the particular condition/variable, for example,
Annual Mean Temperature.

Presence: Each plot tests the climatic variable in relation to the occurrence of
platypus. When the red axis is high (close to 1), it represents a high
presence of platypus occurring in said condition.

Background: When the green plot is closer to 1, it represents areas of Australia's
environment which fits the variable condition, but does not have a high
presence of platypus. If the green plot is close to or at 0, it mean
that Australia has no environments that fit this condition.

Suitability: The blue plot is our models prediction of finding platypus in this
environment, when all other climatic conditions are held at their mean.

### Annual Mean Temperature (C\*100)

```{r}
platypus.bc$response.plots[1]
```

This plot is telling us that platypus are most likely to be found in
environments where the annual mean temperature is below 20$^\circ$C,
with an optimal temperature around 15$^\circ$C.

There are no environments in Australia that have a mean temperature
lower than 10$^\circ$C

### Max Temperature of Warmest Month (C\*100)

```{r}
platypus.bc$response.plots[5]
```

In the warmest month, platypus are not found in any environments where
the temperature excedes 33$^\circ$C. They are found in all environments
where the temperature is lower than 35$^\circ$C

### Mean Temperature of Wettest Quarter (C\*100)

```{r}
platypus.bc$response.plots[8]
```

This plot shows what appears to be two different groups of environments.
One where platypus exist and the wettest quarter of the year's average
temperature is just less than 10$^\circ$C (note than no colder
environments exist in Australia), and one where the wettest quaters mean
temperatures is around 22$^\circ$C.

There are two important things to note here: There is little area of
Australian environmental conditions in between these two groups, and
there is a large area of environment with no platypus, and a much higher
mean temperature during the wettest month.

What is going on here comes down to geography and precipiation. In the
North (North QLD area) where monsoon conditions are present, the wettest
quater of the year is around summer, which is also the hottest. Further
south, in NSW and Vic, the wettest months of the year is around winter,
which is much colder. Finally, the large platypus-absent hot area would
be Australia's arid, dry interior, which has its wettest months in
summer.

This plot is useful in highlighting a flaw in the bioclim model. It is
less complex than a model such as GLM (Generalized Linear Modelling),
which may be able to account for the differing groups regardless, for
the scope of this project, bioclim is the more appropriate due to its
less complex and computationally challenging nature. [Warren, et.al, 2021]

### Other Observations

Platypus can withstand a decently wide range of climates, being found in
Australia's coldest environments (BIO1), but also in the warmer of
regions of Queensland, all the way up to Cape York. Being
aquatic, they are found commonly in climates with decent rainfall around
all months [Hawke et.al, 2019]. Platypus are found more commonly in climates with less
fluctuation in monthly temperatures (BIO3) and less temperature
seasonality (BIO4).


### Interactive Map of BioClim Model

```{r}
# Create an interactive plot to view the model
interactive.plot.enmtools.model(platypus.bc)
```

To see model, untick training points.

Lighter points represent areas with a higher environmental suitability
for platypus habitation. The areas of habitation roughly follow the
Great Dividing Range/East Coast, where platypus are more suited to the cooler,
wetter east side of the mountian range.

Failings of the model: If we compare our model to the observations of platypus, we can see for example, that tasmania has seen many observations, but is ranked as relatively unsuitable by the model.

My guess, is that this is due to a weighting inaccuracies. Platypus, being aquatic care relatively little about precipitation, as long as rivers do not dry up. 7 of the 19 bioclim variables relate to precipitation, possibly over relying on the importance of rainfall in the model. That being said, the model still satisfies the aim of building a species distribution model. 

## Platypus Suitability Locator

Getting input to find platypus suitability

```{r}
# Current Location to assess platypus suitability
# Inputs require decimalLatLong
inputLatitude <- -35.272240
inputLongitude <- 149.121528
  # Current coordinates set to ANU

suitability_value <- extract(platypus.bc$suitability, cbind(inputLongitude, inputLatitude))
suitability_value <- as.numeric(suitability_value)

cat("Locations suitability value: ", suitability_value)
```

Of course, this is modeled likelihood of a complex living organism, in
an overwhelmingly complex living ecosystem. The model does not take into
account human interference, such as daming, pollution, hunting,
urbanisation and environmental destruction, all of which have been known
to negatively impact platypus presence.

The best way to utelise the suitability function is through relative
comparison. Try out a few locations, and have a look what the function
determines is more suitable out of the options.

## Local River Finder

```{r}
# Current Location nearest river
# Using Haversine equation to find the nearest river:

# Function to calculate Haversine distance
nearestRiver <- function(lat1, lon1, lat2, lon2) {
  # Radius of the Earth in km
  earthRadius <- 6371
  
  # Convert degrees to radians
  lat1 <- lat1 * pi / 180
  lon1 <- lon1 * pi / 180
  lat2 <- lat2 * pi / 180
  lon2 <- lon2 * pi / 180
  
  # Haversine formula
  dlat <- lat2 - lat1
  dlon <- lon2 - lon1
  a <- sin(dlat/2)^2 + cos(lat1) * cos(lat2) * sin(dlon/2)^2
  c <- 2 * atan2(sqrt(a), sqrt(1-a))
  distance <- earthRadius * c
  
  return(distance)
}

# Calculate distances and find the nearest point
distances <- sapply(1:nrow(SortedRivers), function(i) {
  nearestRiver(inputLatitude, inputLongitude, SortedRivers$decimalLatitude[i], SortedRivers$decimalLongitude[i])
})

# Find the index of the minimum distance
nearest_point_index <- which.min(distances)

# Get the nearest point
nearest_point <- SortedRivers[nearest_point_index, ]

# Print the nearest point and distance
cat("Locations suitability value: ", suitability_value, "\n",
    "Local River:", nearest_point$SUB_NAME, "river", "\n",
    "Latitude:", nearest_point$decimalLatitude, "\n",
    "Longitude:", nearest_point$decimalLongitude, "\n",
    "Distance (in km) to River:", distances[nearest_point_index], "\n")
```

By altering the values of decimalLatitude and decimalLongitude, users can input their locations, and find the quantitative modeled suitability of anywhere in Australia. Assisting in their search for Australia's most elusive aquatic monotreme.

# References

- Booth, T.H., Nix, H.A., Busby, J.R. and Hutchinson, M.F. (2014), bioclim: the first species distribution modelling package, its early applications and relevance to most current MaxEnt studies. Diversity Distrib., 20: 1-9. https://doi.org/10.1111/ddi.12144

- Gilad Bino, Richard T Kingsford, Michael Archer, Joanne H Connolly, Jenna Day, Kimberly Dias, David Goldney, Jaime Gongora, , The platypus: evolutionary history, biology, and an uncertain future, Journal of Mammalogy, Volume 100, Issue 2, 24 April 2019, Pages 308–327, https://doi.org/10.1093/jmammal/gyz058

- Warren, D.L., Matzke, N.J., Cardillo, M., Baumgartner, J.B., Beaumont, L.J., Turelli, M., Glor, R.E., Huron, N.A., Simões, M., Iglesias, T.L., Piquet, J.C. and Dinnage, R. (2021), ENMTools 1.0: an R package for comparative ecological biogeography. Ecography, 44: 504-511. https://doi.org/10.1111/ecog.05485

- Tahneal Hawke, Gilad Bino, Richard T. Kingsford, A silent demise: Historical insights into population changes of the iconic platypus (Ornithorhynchus anatinus), Global Ecology and Conservation, Volume 20, 2019, ISSN 2351-9894, https://doi.org/10.1016/j.gecco.2019.e00720.

